Approximation as Differentiation

### 2025-07-10 main/conclusion.tex
[ tex ] some future work

@@

\paragraph{More Approximation Domains.} The intervals example is
interesting, but is it useful, and are there more examples like this?
One route might be to follow \cite{edalat-heckmann98} and explore
whether metric spaces (which already provide a native notion of
approximation) can be embedded in $\Fam(\LatGal)$.

\paragraph{Alternative Higher-Order Models}

1. $\Fam(\PSh_{\CMon}(\LatGal))$

2. Concrete sheaves over $\Fam(\LatGal)$ analogous to Diffeological spaces?

\paragraph{Categorical Models of Differentation}

\paragraph{Recursion and Partiality.}
1. Despite using stable domain theory as our starting point, we have ignored

2. Conjecture that the category with $(X : \namedcat{Cpo}, \partial X : X \to \CLatGal)$, where $\CLatGal$ is the category of complete lattices and Galois connections is the right setting. Would need to investigate the connections with dI-domains, bidomains, and bistable bidomains \cite{laird07}.

@@

Lists are the only recursive data type we provided in our source language, so important future work is
supporting general inductive and coinductive types. \citet{nunes2023} support automatic differentiation for
datatypes defined as the least or greatest fixed points of $\mu\nu$-polynomial functors; we could potentially
adopt a similar approach. Full inductive types would allow us to embed an interpreter for untyped lambda
calculus; combined with the CBN monadic translation described in \secref{todo} which uniformly inserts
approximation points, we should be able to obtain the program slicing behaviour of earlier \GPS work ``for
free''. Coinductive types (e.g.~streams) present additional challenges, especially for defining
join-preserving backward maps, but also open the door to slicing (finite prefixes of) infinite data sources,
with some likely relationship to the problem of dealing with partial or non-terminating computations.

>>>>

We have presented a semantic version of \GPS, inspired by connections to differentiable programming and
automatic differentation, and shown that it can be used to interpret an expressive higher-order language
suitable for writing simple queries and data manipulation. Our model elucidates some of the decisions
implicitly taken in previous works on \GPS (\secref{cbn-translation}), and reveals new applications such as
approximation by intervals \secref{interval-approx}. Our categorical approach admits a modular construction of
our model, and the use of general theorems, such as Fiore and Simpson's \thmref{glr-definability}, to prove
properties of the interpretation.

@@

\paragraph{Refinement of the model} As we discussed in \remref{further-structure}, there are possible ways
that the model we have proposed here could be refined to both remove nonsensical elements of the model, and to
augment the model with enough power to prove additional properties such as ``functions are insenstive to
unused inputs''. We are also planning to explore more examples of approximation along the lines of the
intervals example in \secref{interval-approx}. One route might be to follow \citet{edalat-heckmann98}'s
embedding of metric spaces in Scott domains and explore whether metric spaces (which already provide a native
notion of approximation) can be embedded in $\Fam(\LatGal)$.

\paragraph{Categorical Models of Differentation} We have already identified \conref{tangent-stable-fns} and
\conref{tangent-fam} as potential connections to categorical models of differentation. It would also be
interesting to discover any possible links to Differential Linear Logic
\cite{ehrhard_differential_2006}. Differential Linear Logic and the Dialectica translation have been used to
give an account of reverse differentiation by \citet{kerjean-pedrot2024} that is similar to the CHAD approach
of \citet{vakar22}.

@@

Lists are the only recursive data type we provided in our source language, so important future work is
supporting general inductive and coinductive types. \citet{nunes2023} support automatic differentiation for
datatypes defined as the least or greatest fixed points of $\mu\nu$-polynomial functors; we could potentially
adopt a similar approach. Full inductive types would allow us to embed an interpreter for a small language;
combined with the CBN monadic translation described in \secref{cbn-translation} which uniformly inserts
approximation points, we should be able to obtain the program slicing behaviour of earlier \GPS work ``for
free''. Coinductive types (e.g.~streams) present additional challenges, especially for defining
join-preserving backward maps, but also open the door to slicing (finite prefixes of) infinite data sources,
with some likely relationship to the problem of dealing with partial or non-terminating computations.

\paragraph{Recursion and Partiality} This work has only examined the case for total programs. Even though we
took stable domain theory as our starting point, we did not make any use of directed completeness or similar
properties of domains. We expect that an account of recursion in an extension of the framework discussed so
far would likely involve families of bounded lattices indexed by DCPOs, where the order of the DCPO would be
reflected in embedding-projection relationships between the lattices. There are also other refinements of
stable domain theory such as \citet{berry79}'s bidomains and \citet{laird07} bistable biorders that separate
the extensional and stable orders on the same set, in a way that might be similar to
\exref{intervals-and-maxima-elements}.

### 2025-07-10 main/models.tex
[ tex ] some future work

We have constructed the category
$\Fam(\MeetSLat \times \JoinSLat^\op)$, which is a cartesian closed
category with all coproducts, enough to interpret the total
higher-order language we define in the next section with primitive
types and operations defined in $\Fam(\LatGal)$. However, we are not
guaranteed by construction that at first-order type, the
interpretations are in fact Galois connections. We will rectify this
in \secref{definability} using a logical relations construction.

@@

\begin{conjecture}
  (1) The category $\Fam(\MeetSLat)$ is a Tangent category, with the
  tangent bundle
  $T(X, \partial X) = (\Sigma_{x : X}. \partial X(x), \lambda (x,
  \delta x).\, \downset{\delta x})$. (2) The category $\Fam(\LatGal)$
  is a reverse Tangent category with the analogous definition of
  tangent bundle.
\end{conjecture}

>>>>

We have seen that the category $\Fam(\MeetSLat \times \JoinSLat^\op)$,
which is a cartesian closed category with all coproducts, is enough to
interpret the total higher-order language we define in the next
section with primitive types and operations defined in
$\Fam(\LatGal)$. However, we are not guaranteed by construction that
at first-order type, the interpretations are in fact Galois
connections. We will rectify this in \secref{definability} using a
logical relations construction.

@@

\begin{conjecture}
  \label{con:tangent-fam}
  (1) The category $\Fam(\MeetSLat)$ is a Tangent category, with the
  tangent bundle
  $T(X, \partial X) = (\Sigma_{x : X}. \partial X(x), \lambda (x,
  \delta x).\, \downset{\delta x})$. (2) The category $\Fam(\LatGal)$
  is a reverse Tangent category with the analogous definition of
  tangent bundle.
\end{conjecture}

### 2025-07-10 main/related-work.tex
[ tex ] some future work

\emph{Tangent categories}, due originally to \citet{rosický84} and developed by \citet{cockett14,cockett18},
provide an abstract categorical framework for reasoning about differentiation, inspired by the structure of
the tangent bundle in differential geometry. In a tangent category, each object $X$ is equipped with a tangent
bundle $T(X)$, and each morphism $f: X \to Y$ has a corresponding differential map $T(f): T(X) \to T(Y)$
satisfying axioms analogous to the chain rule and linearity of differentiation. Tangent categories generalise
Cartesian differential categories \cite{cdcs}, which model differentiation over cartesian closed categories
using a syntactic derivative operator. The motivation behind tangent categories is to \roly{Bob to say
something here..}

\subsection{Lens Categories}

\cite{spivak19}

>>>>

\emph{Tangent categories}, due originally to \citet{rosický84} and developed by \citet{cockett14,cockett18},
provide an abstract categorical framework for reasoning about differentiation, inspired by the structure of
the tangent bundle in differential geometry. In a tangent category, each object $X$ is equipped with a tangent
bundle $T(X)$, and each morphism $f: X \to Y$ has a corresponding differential map $T(f): T(X) \to T(Y)$
satisfying axioms analogous to the chain rule and linearity of differentiation. Tangent categories generalise
Cartesian differential categories \cite{cdcs}, which model differentiation over cartesian closed categories
using a syntactic derivative operator. Reverse Tangent categories \cite{reverse-tangents} further axiomatise
the existence of reverse derivatives. In \conref{tangent-stable-fns} and \conref{tangent-fam}, we have
conjectured that the categories we have identified in this paper as models of \GPS are Tangent categories. As
well as providing some substantiation to the claim that \GPS is a form of differentiation, we should be able
to apply general results about Tangent categories to our setting. % BOB: such as?

% \subsection{Differential Linear Logic}

% Differential Linear Logic \cite{dill} is an extension of Linear Logic with a derivative operator that acts
% like

% \subsection{Lens Categories}

% \cite{spivak19}
